---
title: "Interactive model builder in fmri.pipeline"
author: "Michael Hallquist"
date: "11 May 2026"
output:
  html_document:
    code_folding: show
    df_print: kable
    mathjax: default
    number_sections: yes
    theme: spacelab
    toc: yes
    toc_depth: 4
  pdf_document:
    code_folding: show
    df_print: kable
    number_sections: yes
    toc: yes
    toc_depth: 4
vignette: >
  %\VignetteIndexEntry{Interactive model builder in fmri.pipeline}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

# Purpose

`fmri.pipeline` can build Level 1, Level 2, and Level 3 model specifications interactively. The interactive builder is useful when you are learning the package, translating a conceptual model into pipeline objects, or creating a first version of a model specification that you will later export to YAML.

The recommended workflow is:

1. Create a `gpa` object with `setup_glm_pipeline()`.
2. Run the interactive builders once.
3. Export the resulting model specification with `export_glm_config()`.
4. Review and edit the YAML file.
5. Use the YAML file for future reproducible runs.

```{r, eval=FALSE}
gpa <- setup_glm_pipeline(
  analysis_name = "clock_glm",
  scheduler = "slurm",
  output_directory = "/proj/lab/analysis",
  trial_data = trial_df,
  run_data = run_df,
  subject_data = subject_df,
  tr = 0.9,
  l1_models = NULL,
  l2_models = NULL,
  l3_models = NULL
)

gpa <- build_l1_models(gpa)
gpa <- build_l2_models(gpa)
gpa <- build_l3_models(gpa)

export_glm_config(gpa, file = "clock_glm_spec.yaml")
```

You can also request prompts during setup:

```{r, eval=FALSE}
gpa <- setup_glm_pipeline(
  analysis_name = "clock_glm",
  trial_data = trial_df,
  run_data = run_df,
  subject_data = subject_df,
  tr = 0.9,
  l1_models = "prompt",
  l2_models = "prompt",
  l3_models = "prompt"
)
```

For long-term use, prefer the first pattern because it separates data/setup choices from model construction and makes the resulting model file easier to version-control.

# Before Starting

The builders rely on the input data stored in `gpa`:

- `build_l1_models()` uses `gpa$trial_data`, and optionally `gpa$ppi_data`.
- `build_l2_models()` uses a Level 2 modeling frame composed from run-level and session-level information.
- `build_l3_models()` uses `gpa$subject_data`.

Before entering the builder, check that the relevant columns are present and have the intended type:

```{r, eval=FALSE}
names(gpa$trial_data)
names(gpa$run_data)
names(gpa$subject_data)

summary(gpa$trial_data[, c("cue_onset", "cue_duration", "condition", "prediction_error")])
summary(gpa$run_data[, c("id", "session", "run_number")])
summary(gpa$subject_data[, c("id", "session", "age", "group")])
```

Pay particular attention to these issues:

- Onsets and durations should be in seconds from the start of each run.
- Trial-level parametric modulators should be numeric.
- Categorical variables should be character or factor if they are truly categorical.
- Integer variables at higher levels may be offered for factor conversion. Convert them when values are category labels rather than numeric trends.
- `subject_data` should have one row per `id` and `session`.

# Level 1 Builder

Level 1 models define run-level regressors. The builder has eight stages:

1. Choose possible onset columns.
2. Choose possible duration columns.
3. Choose possible ISI/ITI columns.
4. Choose possible parametric value columns.
5. Choose possible within-subject factor columns.
6. Build events from onset, duration, and optional ISI/ITI.
7. Build signals from events, amplitudes, HRF normalization, and advanced regressor settings.
8. Build models from signals and contrasts.

Run it with:

```{r, eval=FALSE}
gpa <- build_l1_models(gpa)
```

You can pre-seed column detection if your names are unusual:

```{r, eval=FALSE}
gpa <- build_l1_models(
  gpa,
  onset_cols = c("cue_time", "feedback_time"),
  duration_cols = c("cue_dur", "feedback_dur"),
  value_cols = c("prediction_error", "value"),
  isi_cols = "iti"
)
```

## Stage 1: Onset Columns

The builder first looks for likely event onset columns using the default regex `.*(onset|time).*`. It asks:

```text
Do you want to add these columns to possible onsets?
```

Answer **Yes** if the detected columns are valid event onsets. Answer **No** if the regex captured unrelated time variables, then use the follow-up menu to add or remove columns manually.

Decision points:

- Include columns that mark the start of task events you may want to model.
- Do not include scan metadata, response time columns, or absolute clock timestamps unless they are true event onsets relative to scan start.
- It is fine to include extra onset columns here. You choose which ones become events later.

Example columns:

```text
cue_onset
choice_onset
feedback_onset
```

## Stage 2: Duration Columns

The builder searches for likely duration columns using the default regex `.*duration.*`. It asks whether to add detected duration columns, then allows manual editing.

Use duration columns when duration varies by trial:

```text
cue_duration
choice_duration
feedback_duration
```

If an event has a fixed duration, you do not need a duration column. During event setup, you can choose **Specify fixed duration** and enter a value in seconds.

Decision points:

- Use `0` for impulse-like events if that is scientifically intended.
- Use a fixed value for constant-duration events.
- Use a column for reaction-time or event-duration models.
- Values above 50 seconds trigger a warning because they often indicate milliseconds were provided by mistake.

## Stage 3: ISI/ITI Columns

The builder searches for likely interstimulus or intertrial interval columns using the default regex `^(iti|isi).*`. ISI/ITI columns are optional.

The prompt is:

```text
Do you want to add these columns to possible ISI/ITIs?
```

Use ISI/ITI columns when you want the event object to know what interval follows the event. This can be useful for run truncation logic and for keeping event definitions explicit.

Decision points:

- Include columns such as `iti_duration`, `isi`, or `jitter`.
- Do not include event onset columns here.
- If you do not need ISI/ITI information, leave this set empty.

## Stage 4: Parametric Value Columns

Parametric value columns are continuous amplitudes that can modulate a signal before convolution. The builder restricts choices to numeric or integer-like columns and asks whether to add selected columns.

Examples:

```text
prediction_error
expected_value
reaction_time
uncertainty
```

Decision points:

- Include only columns that should be interpretable as continuous trial-level amplitudes.
- Avoid categorical variables here. Use within-subject factors for categorical trial conditions.
- By default, design-matrix generation centers parametric values (`additional$bdm_args$center_values = TRUE`). This usually helps interpret parametric effects alongside unit-height event regressors.

## Stage 5: Within-Subject Factor Columns

Within-subject factors are trial-level categorical or discrete predictors that can modulate a signal. Examples include condition labels, feedback type, stimulus category, or drug condition within subject.

Examples:

```text
condition
feedback_type
stimulus_category
```

Decision points:

- Choose variables that define trial classes within a run.
- Use within-subject factors when you want separate regressors or contrasts for categories of the same event.
- Do not use subject-level predictors here; those belong in Level 3.

## Stage 6: Events

Events define when something happened. Each event has:

- an event name
- one onset column
- a duration, either fixed or from a column
- optional ISI/ITI, either fixed or from a column

The event menu offers:

```text
Add event
Delete event
Done with event specification
```

When adding an event, the builder asks:

```text
Enter the event name:
Choose event onset
Choose event duration
Do you want to specify an ITI or ISI that follows this event?
```

Example event choices:

```text
Event name: cue
Onset: cue_onset
Duration: cue_duration
ISI/ITI: iti_duration

Event name: feedback
Onset: feedback_onset
Duration: 0.9
ISI/ITI: none
```

Decision points:

- Use short, stable event names. These names appear later in YAML and signal definitions.
- Multiple events can come from the same trial row.
- Event names and signal names can be the same, but they represent different concepts: events are timing definitions; signals are regressors.

Equivalent YAML:

```yaml
events:
  cue:
    onset: cue_onset
    duration: cue_duration
    isi: iti_duration

  feedback:
    onset: feedback_onset
    duration: 0.9
```

## Stage 7: Signals

Signals define regressors. A signal combines an event, a pre-convolution amplitude, HRF normalization, and advanced options.

The signal menu offers:

```text
Add signal
Modify signal
Delete signal
Done with signal setup
```

When adding a signal, the builder asks about:

1. Signal name
2. Event alignment
3. Trial subsetting
4. Regressor value before convolution
5. Within-subject factor modulation
6. HRF normalization
7. Advanced options

### Signal Name

The signal name is the user-facing label for the regressor or regressor family.

Examples:

```text
cue
feedback
pe_feedback
feedback_by_condition
```

Use names that communicate both the event and the effect being estimated.

### Event Alignment

The builder asks:

```text
With which event is this signal aligned?
```

Choose the event whose onset and duration should be used for this signal.

Example:

- A unit-height feedback regressor aligns with the `feedback` event.
- A prediction-error regressor at feedback also aligns with the `feedback` event.

### Trial Subsetting

The builder asks:

```text
Only model this signal for specific trials?
```

If you answer **Yes**, you enter an R expression evaluated against `trial_data`. The expression must return `TRUE` or `FALSE` for each trial row.

Examples:

```r
reaction_time < 4
accuracy == 1
condition == "reward"
reaction_time < 4 & accuracy == 1
```

Decision points:

- Use trial subsetting when a regressor should only exist for a subset of trials.
- The builder rejects expressions that retain zero trials.
- Keep the expression simple enough to audit later in exported YAML.

Equivalent YAML:

```yaml
signals:
  correct_feedback:
    event: feedback
    trial_subset_expression: accuracy == 1
    value_fixed: 1
    normalization: none
```

### Regressor Value

The builder asks:

```text
What should be the value of regressor (pre-convolution)?
```

Options:

- **Unit height (1.0)**: a standard event occurrence regressor.
- **Other fixed value**: a constant amplitude other than 1.
- **Parametric modulator**: a trial-level numeric column selected earlier.

Decision points:

- Use unit height for event occurrence.
- Use a parametric modulator for trialwise continuous effects such as prediction error.
- For a parametric effect, usually include the corresponding unit-height event signal in the same model so the parametric regressor is interpreted as modulation around event occurrence.

Equivalent YAML:

```yaml
signals:
  feedback:
    event: feedback
    value_fixed: 1
    normalization: none

  pe_feedback:
    event: feedback
    parametric_modulator: prediction_error
    normalization: evtmax_1
```

### Within-Subject Factor Modulation

If within-subject factor columns were selected earlier, the builder asks:

```text
Is this signal modulated by one or more within-subject factors?
```

If you answer **Yes**, you enter an R model formula using selected within-subject factors:

```r
~ condition
~ feedback_type
~ condition + feedback_type
~ condition * feedback_type
```

The pipeline removes the intercept internally for the within-subject model so that category-specific signal columns are easier to contrast.

Decision points:

- Use a within-subject factor when you need separate regressors by condition for the same event.
- Use `~ condition` for one factor.
- Use `~ condition * feedback_type` when interaction cells are scientifically meaningful and sufficiently populated.
- Avoid very sparse factorial models; empty or rare cells can lead to aliased regressors and fragile contrasts.

Equivalent YAML:

```yaml
signals:
  feedback_by_condition:
    event: feedback
    value_fixed: 1
    normalization: none
    wi_factors:
      - condition
    wi_formula: ~ condition - 1
```

### HRF Normalization

The builder asks:

```text
How should the HRF be normalized in convolution?
```

Options:

- `none`: no HRF rescaling before convolution.
- `evtmax_1`: event-wise normalization so each event's HRF peak is 1.0 regardless of duration.
- `durmax_1`: duration-sensitive normalization that approaches a peak of 1.0 for long events.

Practical guidance:

- Use `none` for ordinary unit-height event regressors when the default scaling is desired.
- Use `evtmax_1` for many parametric modulators so the amplitude column has a more direct interpretation.
- Use `durmax_1` when duration is itself part of the modeled response magnitude.

### Advanced Signal Options

For a new signal, the builder prints the defaults:

```text
No temporal derivative
Demean convolved signal
No beta series
No time series multiplier (PPI)
```

It then asks:

```text
Accept default advanced options for this signal?
```

If you answer **No**, it asks:

- `Add temporal derivative?`
- `Demean signal post-convolution?`
- `Generate beta series for this signal (one regressor per trial)?`
- `Multiply this signal against a physiological regressor for PPI analysis?`

Decision points:

- Temporal derivatives can help absorb modest HRF timing differences, but add columns and complicate contrasts.
- Post-convolution demeaning is the default and is usually appropriate.
- Beta series creates one regressor per trial and should be used only when you need trialwise estimates, such as some connectivity workflows.
- PPI multiplication requires `ppi_data`; otherwise the option is disabled.

## Common L1 Signal Patterns

These patterns cover most first-pass task models. They assume that `events:` already contains timing definitions such as `cue` and `feedback`.

### Event Occurrence

Use a unit-height signal when the question is whether BOLD activity differs from baseline during an event.

```yaml
signals:
  cue:
    event: cue
    value_fixed: 1
    normalization: none
```

This creates one convolved regressor for cue occurrence. It is the safest starting point for most events.

### Event Plus Parametric Modulator

Use paired occurrence and parametric signals when a trialwise continuous value should explain variation around the event response.

```yaml
signals:
  feedback:
    event: feedback
    value_fixed: 1
    normalization: none

  pe_feedback:
    event: feedback
    parametric_modulator: prediction_error
    normalization: evtmax_1

l1_models:
  feedback_pe:
    signals:
      - feedback
      - pe_feedback
    contrasts:
      diagonal: yes
```

The unit-height `feedback` signal captures the average feedback response. The `pe_feedback` signal captures whether the response varies with `prediction_error`. Omitting the occurrence signal can make the parametric effect absorb the event mean, which is usually not what you want.

### Condition-Specific Event Regressors

Use within-subject factor modulation when each condition should have its own event regressor.

```yaml
signals:
  feedback_by_condition:
    event: feedback
    value_fixed: 1
    normalization: none
    wi_factors:
      - condition
    wi_formula: ~ condition - 1

l1_models:
  feedback_condition:
    signals:
      - feedback_by_condition
    contrasts:
      diagonal: yes
      wi_contrasts:
        feedback_by_condition:
          diagonal: yes
          pairwise_diffs: condition
```

This is appropriate for categorical trial conditions such as reward/loss, correct/error, or stimulus type. Check that every run has enough trials in each modeled condition; sparse cells can create unstable or aliased regressors.

### Trial-Subset Regressor

Use a trial subset when a regressor should only be present for trials satisfying a clear rule.

```yaml
signals:
  fast_correct_feedback:
    event: feedback
    trial_subset_expression: reaction_time < 1.5 & accuracy == 1
    value_fixed: 1
    normalization: none
```

This creates a feedback occurrence regressor only for fast correct trials. Keep subset expressions simple and inspect the number of retained trials by subject/run before trusting the model.

### Beta-Series Signal

Use beta series only when trialwise estimates are needed.

```yaml
signals:
  feedback_beta_series:
    event: feedback
    value_fixed: 1
    normalization: none
    beta_series: yes
```

This creates one regressor per trial for the signal. It can be useful for trialwise connectivity or prediction workflows, but it greatly increases model size and is not a default choice for ordinary activation GLMs.

## Stage 8: L1 Models and Contrasts

An L1 model selects a set of signals and creates contrasts over the resulting run-level regressors.

The L1 model menu offers:

```text
Add model
Modify model
Delete model
Done with l1 model setup
```

When adding a model, the builder asks:

```text
Enter the model name:
Choose all signals to include in this model
```

Then it calls the contrast builder.

Decision points:

- Start with one simple L1 model before adding variants.
- Include unit-height event regressors alongside parametric modulators.
- Do not include redundant signals unless you have a clear reason; collinearity will propagate upward.

Equivalent YAML:

```yaml
l1_models:
  task:
    signals:
      - cue
      - feedback
      - pe_feedback
    contrasts:
      diagonal: yes
```

# Contrast Builder

The contrast builder is shared by L1, L2, L3, and within-subject factor submodels. It creates the contrast matrix stored in the model object.

## Diagonal Contrasts

For ordinary L1 models without within-subject factor contrasts, diagonal contrasts are added automatically. For L2 and L3 models, the builder asks:

```text
Do you want to include diagonal contrasts for each regressor?
```

Diagonal contrasts create one contrast per design-matrix column. In FSL terms, these become COPEs for individual explanatory variables.

Decision points:

- Usually answer **Yes**.
- For intercept-only L2/L3 models, the pipeline automatically adds the intercept contrast.
- You may delete unwanted diagonal contrasts in the contrast editor.

## Emmeans Weighting

For models with a fitted `lm` object, usually L2, L3, or L1 within-subject factor models, the builder asks:

```text
How should contrasts be computed by emmeans?
```

Options:

- `equal`: all cells weighted equally.
- `cells`: weights use observed cell frequencies. This is the package's usual recommendation for unbalanced designs.
- `proportional`: weights use frequencies of factor combinations.
- `flat`: weights observed cells while ignoring empty cells.

Decision points:

- Use `cells` for most unbalanced fMRI designs.
- Use `equal` when each condition should contribute equally by design, regardless of observed frequencies.
- Be cautious with empty cells; they can make some emmeans contrasts non-estimable.

## Condition Means

For each categorical predictor, the builder asks:

```text
Do you want to include model-predicted means for <factor>?
```

Condition means estimate marginal means for each factor level.

Use them when you want maps for each category, such as `groupcontrol` and `grouppatient`, or each session label.

## Pairwise Differences

For categorical predictors and categorical interactions, the builder asks:

```text
Do you want to include pairwise differences for <factor>?
```

Pairwise differences estimate contrasts between factor levels, such as `patient - control` or `followup - baseline`.

Use them when the scientific question is a category comparison rather than a category mean.

## Cell Means

When a model has more than one categorical factor, the builder asks whether to include cell means across the factors.

Use cell means when the full factorial combinations are meaningful, such as `group` by `session_label`.

## Overall Response

The builder asks:

```text
Do you want to include the overall average response?
```

This creates a contrast for the model-predicted average response across predictors. It is often useful for group-level summaries and quality checks.

## Simple Slopes

If the model includes interactions between continuous and categorical predictors, the builder may ask whether to include model-predicted simple slopes across factor levels.

Use this when the model includes terms such as:

```r
~ age * group
```

and you want an age slope estimated separately within each group.

## Custom Contrast Editor

After automatic contrasts are generated, the contrast editor offers:

```text
Add contrast
Show contrasts
Delete contrast
Done with contrast setup
```

For custom contrasts, you can enter either a numeric vector:

```text
1 0 -1 0
```

or named coefficients:

```text
condition_reward = 1 condition_loss = -1
```

Named coefficients are safer because they do not depend on remembering column order.

# Level 2 Builder

Level 2 models combine run-level estimates within subject or within subject-session.

Run it with:

```{r, eval=FALSE}
gpa <- build_l2_models(gpa)
```

The L2 menu offers:

```text
Add model
Modify model
Delete model
Done with second-level (subject) model setup
```

When adding a model, the builder asks about:

1. Model name
2. Formula entry versus model-builder walkthrough
3. Variable transformations and factor reference levels
4. Session-combination scope
5. Contrasts

## Formula or Model Builder

The builder asks:

```text
Do you want to specify the model formula or walk through the model builder?
```

Choose **Specify formula** when you know the R formula:

```r
~ 1
~ condition
~ difficulty + run_number
~ condition * difficulty
```

Choose **Model builder** for a simple additive model. The walkthrough lets you select variables and creates a formula like:

```r
~ condition + difficulty
```

Decision points:

- Use `~ 1` for a simple run-average model.
- Use formula entry for interactions.
- Use the model-builder walkthrough for additive models only.

## Continuous Covariate Transformations

For continuous variables, the builder asks:

```text
Transform <variable>?
```

Options:

- Mean center
- Standardize/z-score
- Subtract minimum
- Subtract maximum
- No transformation

Decision points:

- Mean-center most continuous covariates unless zero is already meaningful.
- Standardize when effect sizes per standard deviation are easier to interpret.
- Subtract minimum or maximum when a boundary value is scientifically meaningful.
- Transformations are recomputed after dropped runs or subjects are removed.

## Factor Reference Levels

For factor variables, the builder asks:

```text
Choose the reference level for the <factor> factor.
```

The default is the most frequent level. The reference level affects treatment-coded regressors and the interpretation of intercepts and some contrasts.

Decision points:

- Use the control or baseline category as reference when available.
- For session factors, baseline is usually the clearest reference.
- Pairwise differences from `emmeans` are less dependent on the reference level than raw model coefficients.

## L2 Session Scope

If the data contain multiple sessions, the L2 builder asks:

```text
How should this L2 model combine sessions?
```

Options:

- **Within-session**: `l2_scope = "id_session"`. Runs are combined separately within each subject-session.
- **Across-session**: `l2_scope = "id"`. Runs from all sessions are stacked into one subject-level model.

Decision points:

- Use `id_session` for most longitudinal analyses. It preserves session-level outputs for Level 3.
- Use `id` only when pooling sessions within subject is scientifically intended.
- If you need AFNI `3dLMEr` at Level 3, use `id_session`.
- Between-session contrasts at L2 require `id`, but that choice pools sessions in a way that is not appropriate for many longitudinal designs.

Equivalent YAML:

```yaml
l2_models:
  run_average:
    level: 2
    model_formula: ~ 1
    l2_scope: id_session
    contrasts:
      diagonal: yes
      overall_response: yes
```

# Level 3 Builder

Level 3 models define group/sample-level inference. The implementation uses the same builder as Level 2, with Level 3-specific questions.

Run it with:

```{r, eval=FALSE}
gpa <- build_l3_models(gpa)
```

The L3 menu offers:

```text
Add model
Modify model
Delete model
Done with third-level (sample) model setup
```

When adding a model, the builder asks about:

1. Model name
2. Formula entry versus model-builder walkthrough
3. Variable transformations and factor reference levels
4. L3 session input mode
5. `3dLMEr` random effects, when applicable
6. Contrasts
7. FSL outlier deweighting

## L3 Formula

Common L3 formulas:

```r
~ 1
~ group
~ age
~ group + age
~ group * session_label
```

Decision points:

- Use `~ 1` for the sample mean of lower-level contrasts.
- Add covariates such as `age` or motion summaries when they are part of the planned group model.
- Use interaction terms only when the design and sample size support them.

## L3 Session Input Mode

If there are multiple sessions, the builder asks:

```text
How should this L3 model use sessions?
```

Options:

- `per_session`: fit separate per-session L3 inputs. Requires `l2_scope = "id_session"`.
- `pooled_sessions_subject_ev`: pool sessions in FSL with subject indicator EVs. Requires `l2_scope = "id_session"`.
- `subject_rows`: use one row per subject input. Requires `l2_scope = "id"`.
- `3dlmer`: use AFNI `3dLMEr` for a longitudinal mixed-effects L3 model. Requires `l2_scope = "id_session"`.

Decision points:

- Use `per_session` for separate baseline/follow-up group maps.
- Use `pooled_sessions_subject_ev` for FSL-native longitudinal models with subject EVs.
- Use `3dlmer` when you need mixed-effects longitudinal modeling, uneven visit spacing, missing sessions, continuous time, or random effects.
- Avoid `subject_rows` unless the L2 model intentionally pooled sessions with `l2_scope = "id"`.

Equivalent YAML for an FSL per-session group model:

```yaml
l3_models:
  group_mean:
    level: 3
    model_formula: ~ 1
    l3_input_mode: per_session
    contrasts:
      diagonal: yes
      overall_response: yes
    fsl_outlier_deweighting: no
```

## AFNI 3dLMEr Random Effects

If you choose `l3_input_mode = "3dlmer"`, the builder asks:

```text
Enter the random-effects specification:
```

Examples:

```r
(1 | Subj)
(1 + session_label | Subj)
```

The grouping variable `Subj` is recommended for AFNI compatibility. The builder combines your fixed-effects formula and random-effects specification into a full `3dLMEr` formula.

Equivalent YAML:

```yaml
l3_models:
  longitudinal_lmer:
    level: 3
    model_formula: ~ session_label
    l3_input_mode: 3dlmer
    execution_backend: afni
    producer_backend: fsl
    random_effects: "(1 | Subj)"
    contrasts:
      cond_means: session_label
      pairwise_diffs: session_label
      overall_response: yes
```

## FSL Outlier Deweighting

For L3 models, the builder asks:

```text
Turn on outlier deweighting?
```

This controls FSL robust outlier deweighting for mixed-effects Level 3 analyses.

Decision points:

- Outlier deweighting is often helpful for group analyses.
- Disable it if FLAME reports excessive outliers for otherwise legitimate inputs.
- This setting is FSL-specific; AFNI `3dLMEr` uses AFNI's modeling path.

# Modifying Existing Models

Each builder can modify an existing model set. If `gpa$l1_models`, `gpa$l2_models`, or `gpa$l3_models` already exists, the menu offers options to modify or delete components.

Use modification carefully:

- Deleting an onset, duration, event, or signal does not always cascade-delete dependent objects.
- If you substantially change early L1 pieces, inspect later events, signals, models, and contrasts.
- For major changes, it is often cleaner to edit the exported YAML and rebuild from the spec file.

Example rebuild:

```{r, eval=FALSE}
gpa$l1_models <- NULL
gpa$l2_models <- NULL
gpa$l3_models <- NULL

gpa <- build_l1_models(gpa, from_spec_file = "clock_glm_spec.yaml")
gpa <- build_l2_models(gpa, from_spec_file = "clock_glm_spec.yaml")
gpa <- build_l3_models(gpa, from_spec_file = "clock_glm_spec.yaml")
```

# Practical Patterns

## Simple Multi-Run Task

Use this pattern for a standard task with multiple runs per subject:

```yaml
l2_models:
  run_average:
    level: 2
    model_formula: ~ 1
    l2_scope: id_session
    contrasts:
      diagonal: yes
      overall_response: yes

l3_models:
  group_mean:
    level: 3
    model_formula: ~ 1
    l3_input_mode: per_session
    contrasts:
      diagonal: yes
      overall_response: yes
    fsl_outlier_deweighting: no
```

## Group Difference at L3

Use this when a subject/session-level variable defines groups:

```yaml
l3_models:
  group_difference:
    level: 3
    model_formula: ~ group
    reference_level:
      group: control
    l3_input_mode: per_session
    contrasts:
      diagonal: yes
      cond_means: group
      pairwise_diffs: group
      overall_response: yes
    fsl_outlier_deweighting: no
```

## Trial Condition at L1

Use within-subject factor modulation when the same event should produce condition-specific regressors:

```yaml
signals:
  feedback_by_condition:
    event: feedback
    value_fixed: 1
    normalization: none
    wi_factors:
      - condition
    wi_formula: ~ condition - 1

l1_models:
  feedback_condition:
    signals:
      - feedback_by_condition
    contrasts:
      diagonal: yes
      wi_contrasts:
        feedback_by_condition:
          diagonal: yes
          pairwise_diffs: condition
```

# Final Checks

After using the builders, inspect and export the model object:

```{r, eval=FALSE}
summary(gpa)

names(gpa$l1_models$events)
names(gpa$l1_models$signals)
names(gpa$l1_models$models)
names(gpa$l2_models$models)
names(gpa$l3_models$models)

export_glm_config(gpa, file = "glm_spec.yaml")
```

Then rebuild from the exported file in a clean R session:

```{r, eval=FALSE}
gpa$l1_models <- NULL
gpa$l2_models <- NULL
gpa$l3_models <- NULL

gpa <- build_l1_models(gpa, from_spec_file = "glm_spec.yaml")
gpa <- build_l2_models(gpa, from_spec_file = "glm_spec.yaml")
gpa <- build_l3_models(gpa, from_spec_file = "glm_spec.yaml")
```

This verifies that your interactive choices can be reproduced non-interactively before you submit a full pipeline run.
